White Dwarf Mergers on Adaptive Meshes\\ I. Methodology and Code Verification

### 2017-02-25 papers/wdmerger_I/submit2/paper.tex
Commit final submitted version of paper I

\documentclass[twocolumn,numberedappendix]{../aastex6}

@@

\begin{figure*}[ht]
  \centering
  \includegraphics[trim=0.75in 0.45in 0.85in 0.5in, clip, scale = 0.75]{{{plots/kh_t2.0_p3_low_res_collage}}}
  \caption{2D Kelvin-Helmholtz instability test at $t = 2.0$ for the initial 
    conditions given by \autoref{eq:kh_p3_rho} through \autoref{eq:kh_p3_vy},
    which come from McNally et al. (2012).
    The meaning of the rows and columns is the same as in \autoref{fig:kh_p2}.
    \label{fig:kh_p3}}
\end{figure*}

@@

\begin{figure*}[ht]
  \includegraphics[trim = 0in 1.25in 0.25in 1.25in, clip, scale=0.65]{{{plots/kh_p3_high_res_collage}}}
  \caption{Time series of the Kelvin-Helmholtz problem proposed by McNally et al. (2012)
    as the simulation is just starting to go non-linear. The rows represent resolution, 
    where $n$ is the number of grid cells per spatial dimension, and the columns are 
    different snapshots in time.\label{fig:kh_p3_high_res}}
\end{figure*}

\citet{mcnally:2012} published another Kelvin-Helmholtz problem that 
is well-posed in the sense that it converges with resolution and 
is not subject to uncontrollable numerical instabilities. Though they 
were not explicitly interested in the question of Galilean invariance, 
we visit that issue here to see what can be learned. The initial 
conditions for this problem are:
\begin{align}
  \rho &= \begin{dcases} \rho_1 - \rho_m e^{(y-0.25)/\Delta y} & 0.25 > y \geq 0 \\ 
                         \rho_2 + \rho_m e^{(0.25-y)/\Delta y} & 0.5 > y \geq 0.25 \\
                         \rho_2 + \rho_m e^{(y-0.75)/\Delta y} & 0.75 > y \geq 0.5 \\
                         \rho_1 - \rho_m e^{(0.75-y)/\Delta y} & 1 > y \geq 0.75 \end{dcases} \label{eq:kh_p3_rho}
\end{align}
\begin{align}
  v_x &= \begin{dcases} v_1 - v_m e^{(y-0.25)/\Delta y} & 0.25 > y \geq 0 \\
                        v_2 + v_m e^{(0.25-y)/\Delta y} & 0.5 > y \geq 0.25 \\
                        v_2 + v_m e^{(y-0.75)/\Delta y} & 0.75 > y \geq 0.5 \\
                        v_1 - v_m e^{(0.75-y)/\Delta y} & 1 > y \geq 0.75 \end{dcases} \label{eq:kh_p3_vx} \\
  v_y &= w_0\, \text{sin}\left(4\pi x\right). \label{eq:kh_p3_vy}
\end{align}
Here $\Delta y = 0.025$, $w_0 = 0.01$, $v_m = (v_1 - v_2) / 2$, $\rho_m = (\rho_1 - \rho_2) / 2$, 
and the other symbols have the same meaning as above (this means the flow direction
is reversed compared to the original paper, so as to achieve consistency with the 
other simulations presented here). We run this problem at all the same resolutions 
and bulk velocities as the previous two problems. The results for the normal resolutions 
at $t = 2.0$ are displayed in \autoref{fig:kh_p3}. We see a similar pattern as for 
the test proposed by R10: as we get to higher flow speeds we need to have higher 
spatial resolution to compensate for the increased numerical diffusion. The 
qualitative accuracy is much lower for the highest bulk velocities for this problem
than for the previous problems. This is because the amplitude of the instability overall
is smaller than for the previous problems, at least by $t = 2.0$, so it is easier
for numerical diffusion at the shearing layer, caused by the high bulk velocities,
to completely wipe out the instability. Like \citet{robertson:2010} found for
their problem, we find for this problem that the convergence properties are not
substantially affected by altering the perturbation frequency -- the results
show the same qualitative pattern even if we halve this frequency.

@@

\input{plots/sources.table}

@@

\begin{figure}
  \centering
  \includegraphics[scale=0.45]{plots/unequal_gw_rot0}
  \caption{Gravitational wave strain polarizations for the first two orbital periods of an 
           unequal mass system. The curves with markers are the numerical data, while the 
           curves without markers are the analytical results for two point masses.\label{fig:gw_strain}}
\end{figure}

Finally we consider whether the dynamical behavior of the system converges with resolution. 
In \autoref{fig:unequal_spatial_convergence_inertial} we plot the first full orbit for 
the unequal mass system, at three different resolutions in the inertial frame: our default 
resolution of $256^3$ zones, as well as a single level of refinement with a jump by a factor of 
two (effective resolution $512^3$) or a jump by a factor of four (effective resolution $1024^3$). 
It is clear that at the latter resolution (corresponding to physical resolution of 100 km), 
we have achieved convergent behavior. In the rotating frame, the results also show
convergent behavior but the convergence is not as fast with resolution as in the inertial frame; 
see \autoref{fig:unequal_spatial_convergence_rotating}. At the two higher resolutions the white dwarf
distance is qualitatively similar, and both are qualitatively different from the lower resolution. However,
quantitatively the two higher resolution runs are not as similar to each other as the analogous runs in the
inertial frame. Convergence with resolution is slightly slower in the rotating reference frame
because in the rotating reference frame a stable, unchanging circular orbit requires balance between
two forces with opposite sign (the gravitational and centrifugal forces), and slight perturbations from the
circular orbit are amplified by the effect of the Corolis force. In the inertial frame, these numerical instabilities
vanish, but the cost is that there is no centrifugal force to actively maintain the white dwarf distance,
which is why it is much more likely for the orbit to prematurely decay. In either case, these results suggest
at least a minimum resolution of 200 km for getting the dynamics qualitatively right. To put that into context,
consider that the parameter study of \citet{dan:2014} used 40,000 SPH particles per simulation, or (for an equal mass
binary) 20,000 particles per white dwarf. For, say, a $0.9\ \msolar + 0.9\ \msolar$ white dwarf binary on
a $256^3$ zone simulation grid, there are 20,000 zones that fit within a white dwarf. We do not
intend here to directly compare results between the two simulation methods. We limit ourselves to the
observation that at least for grid-based codes, a parameter study such as the ones performed by
\citet{dan:2012} and \citet{dan:2014} would likely not yield qualitatively convergent results if it were to use the
same effective mass resolution. Instead the number of zones inside each star should at least be doubled.

\begin{figure}
  \centering
  \includegraphics[scale=0.45]{plots/spatial_convergence_rot0}
  \caption{Distance between the two white dwarfs in the unequal mass system, for the first orbit.
           The distance is scaled by the initial orbital distance. 
           We plot at three different resolutions, corresponding to the number of 
           effective zones per dimension in the refined regions.
           \label{fig:unequal_spatial_convergence_inertial}}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[scale=0.45]{plots/spatial_convergence_rot1}
  \caption{Distance between the two white dwarfs in the unequal mass system, for the first orbit.
           The distance is scaled by the initial orbital distance. 
           We plot at three different resolutions, corresponding to the number of 
           effective zones per dimension in the refined regions.
           \label{fig:unequal_spatial_convergence_rotating}}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[scale=0.4]{plots/weak_scaling}
  \caption{\castro\ weak scaling test, performed on Blue Waters at 
    NCSA. Each processor had a fixed amount of work, and we increased the 
    number of simulation zones in concert with the number of processors. The 
    solid curve represents perfect weak scaling, while the blue circles show 
    \castro's performance at each processor count. The vertical axis measures 
    the median time per timestep, normalized to this value for the smallest 
    processor count.\label{fig:weak_scaling}}
\end{figure}

To examine the parallel performance of \castro, we performed both
strong scaling and weak scaling tests on the Blue Waters machine at
the National Center for Supercomputing Applications. For the weak
scaling test, whose results are shown in \autoref{fig:weak_scaling},
we ran a uniform grid binary white dwarf simulation for resolutions of
$128^3$ zones through $2048^3$ zones. The number of processors was
scaled with the number of zones so that each processor had the same
amount of work; the smallest test used 8 processors and the largest
used 32,768 (note that the number of processor cores on a Blue Waters
node is twice the number of floating point units on that node). The
test was run for 10 timesteps, with each timestep including two
Poisson solves and a hydrodynamics update (though for a uniform grid
calculation we generally do not need to perform any multigrid
iterations for the first Poisson solve in a timestep, since the
density distribution has not changed since the end of the last
timestep). We disabled plotfile and checkpoint writing, as well as
calculation of diagnostic information (the latter can contribute to a
significant fraction of the run time at large processor counts if
computed every timestep). We computed the median wall time required
per time step for each simulation, and then normalized this to the
median time per timestep for the smallest simulation. We find
excellent weak scaling through 4,096 processors. At the largest run,
the simulation time required is slightly less than 1.5 times the amount
required for the smallest simulation.  This is due entirely to the
increased cost of the multigrid Poisson solve in each timestep and
this cannot be mitigated except by improving communication or
computation efficiency in the multigrid solver. We observe that this
weak scaling behavior with Poisson gravity is a significant
improvement over the results presented in the first \castro\ paper.

\begin{figure}
  \centering
  \includegraphics[scale=0.4]{plots/strong_scaling}
  \caption{\castro\ strong scaling test performed on the Blue Waters machine at
    NCSA. The vertical axis measures the median time per timestep, and the 
    horizontal axis measures the number of processors in the simulation. Data 
    points are normalized to the time per timestep for the smallest number 
    of processors. The green circles show the data for a simulation with one 
    AMR level (a single uniform grid), the blue diamonds show the data for a simulation with two AMR
    levels (one coarse and one fine), while the red circles show the data 
    for a test with three AMR levels (one coarse and two fine). The fine levels 
    increase the resolution only in the regions around the stars. For each case 
    we draw a solid curve representing perfect strong scaling.\label{fig:strong_scaling}}
\end{figure}

@@

The authors thank John Bell, Volker Springel, and Dominic Marcello
for helpful discussions on gravity 
and hydrodynamics issues. We also thank Noel Scudder and Platon Karpov 
for their help with this project, especially related to visualization of the results, 
and Adam Jacobs for his advice on and assistance with 
running on supercomputing resources. Special thanks are given 
to the organizers of the 2015 Caltech Gravitational Wave 
Astrophysics School, and their supporters: the NSF 
under CAREER award PHY-1151197, the Sherman Fairchild 
Foundation, the LIGO Laboratory at Caltech, and Caltech.

@@

\bibliographystyle{../aasjournal}
\bibliography{../refs}

>>>>

\documentclass[twocolumn,numberedappendix]{aastex6}

@@

\begin{figure*}[ht]
  \centering
  \includegraphics[trim=0.75in 0.45in 0.85in 0.5in, clip, scale = 0.75]{{{kh_t2.0_p3_low_res_collage}}}
  \caption{2D Kelvin-Helmholtz instability test at $t = 2.0$ for the initial 
    conditions given by \autoref{eq:kh_p3_rho} through \autoref{eq:kh_p3_vy},
    \textbf{which come from McNally et al. (2012)}.
    The meaning of the rows and columns is the same as in \autoref{fig:kh_p2}.
    \label{fig:kh_p3}}
\end{figure*}

@@

\begin{figure*}[ht]
  \includegraphics[trim = 0in 1.25in 0.25in 1.25in, clip, scale=0.65]{{{kh_p3_high_res_collage}}}
  \caption{Time series of the Kelvin-Helmholtz problem proposed by McNally et al. (2012)
    as the simulation is just starting to go non-linear. The rows represent resolution, 
    where $n$ is the number of grid cells per spatial dimension, and the columns are 
    different snapshots in time.\label{fig:kh_p3_high_res}}
\end{figure*}

\citet{mcnally:2012} published another Kelvin-Helmholtz problem that 
is well-posed in the sense that it converges with resolution and 
is not subject to uncontrollable numerical instabilities. Though they 
were not explicitly interested in the question of Galilean invariance, 
we visit that issue here to see what can be learned. The initial 
conditions for this problem are:
\begin{align}
  \rho &= \begin{dcases} \rho_1 - \rho_m e^{(y-0.25)/\Delta y} & 0.25 > y \geq 0 \\ 
                         \rho_2 + \rho_m e^{(0.25-y)/\Delta y} & 0.5 > y \geq 0.25 \\
                         \rho_2 + \rho_m e^{(y-0.75)/\Delta y} & 0.75 > y \geq 0.5 \\
                         \rho_1 - \rho_m e^{(0.75-y)/\Delta y} & 1 > y \geq 0.75 \end{dcases} \label{eq:kh_p3_rho}
\end{align}
\begin{align}
  v_x &= \begin{dcases} v_1 - v_m e^{(y-0.25)/\Delta y} & 0.25 > y \geq 0 \\
                        v_2 + v_m e^{(0.25-y)/\Delta y} & 0.5 > y \geq 0.25 \\
                        v_2 + v_m e^{(y-0.75)/\Delta y} & 0.75 > y \geq 0.5 \\
                        v_1 - v_m e^{(0.75-y)/\Delta y} & 1 > y \geq 0.75 \end{dcases} \label{eq:kh_p3_vx} \\
  v_y &= w_0\, \text{sin}\left(4\pi x\right). \label{eq:kh_p3_vy}
\end{align}
Here $\Delta y = 0.025$, $w_0 = 0.01$, $v_m = (v_1 - v_2) / 2$, $\rho_m = (\rho_1 - \rho_2) / 2$, 
and the other symbols have the same meaning as above (this means the flow direction
is reversed compared to the original paper, so as to achieve consistency with the 
other simulations presented here). We run this problem at all the same resolutions 
and bulk velocities as the previous two problems. The results for the normal resolutions 
at $t = 2.0$ are displayed in \autoref{fig:kh_p3}. We see a similar pattern as for 
the test proposed by R10: as we get to higher flow speeds we need to have higher 
spatial resolution to compensate for the increased numerical diffusion. \textbf{The 
qualitative accuracy is much lower for the highest bulk velocities for this problem
than for the previous problems. This is because the amplitude of the instability overall
is smaller than for the previous problems, at least by $t = 2.0$, so it is easier
for numerical diffusion at the shearing layer, caused by the high bulk velocities,
to completely wipe out the instability. Like \citet{robertson:2010} found for
their problem, we find for this problem that the convergence properties are not
substantially affected by altering the perturbation frequency -- the results
show the same qualitative pattern even if we halve this frequency.}

@@

\input{sources.table}

@@

\begin{figure}
  \centering
  \includegraphics[scale=0.45]{unequal_gw_rot0}
  \caption{Gravitational wave strain polarizations for the first two orbital periods of an 
           unequal mass system. The curves with markers are the numerical data, while the 
           curves without markers are the analytical results for two point masses.\label{fig:gw_strain}}
\end{figure}

Finally we consider whether the dynamical behavior of the system converges with resolution. 
In \autoref{fig:unequal_spatial_convergence_inertial} we plot the first full orbit for 
the unequal mass system, at three different resolutions in the inertial frame: our default 
resolution of $256^3$ zones, as well as a single level of refinement with a jump by a factor of 
two (effective resolution $512^3$) or a jump by a factor of four (effective resolution $1024^3$). 
It is clear that at the latter resolution (corresponding to physical resolution of 100 km), 
we have achieved convergent behavior. \textbf{In the rotating frame, the results also show
convergent behavior but the convergence is not as fast with resolution as in the inertial frame; 
see \autoref{fig:unequal_spatial_convergence_rotating}. At the two higher resolutions the white dwarf
distance is qualitatively similar, and both are qualitatively different from the lower resolution. However,
quantitatively the two higher resolution runs are not as similar to each other as the analogous runs in the
inertial frame. Convergence with resolution is slightly slower in the rotating reference frame
because in the rotating reference frame a stable, unchanging circular orbit requires balance between
two forces with opposite sign (the gravitational and centrifugal forces), and slight perturbations from the
circular orbit are amplified by the effect of the Corolis force. In the inertial frame, these numerical instabilities
vanish, but the cost is that there is no centrifugal force to actively maintain the white dwarf distance,
which is why it is much more likely for the orbit to prematurely decay. In either case, these results suggest
at least a minimum resolution of 200 km for getting the dynamics qualitatively right. To put that into context,
consider that the parameter study of \citet{dan:2014} used 40,000 SPH particles per simulation, or (for an equal mass
binary) 20,000 particles per white dwarf. For, say, a $0.9\ \msolar + 0.9\ \msolar$ white dwarf binary on
a $256^3$ zone simulation grid, there are 20,000 zones that fit within a white dwarf. We do not
intend here to directly compare results between the two simulation methods. We limit ourselves to the
observation that at least for grid-based codes, a parameter study such as the ones performed by
\citet{dan:2012} and \citet{dan:2014} would likely not yield qualitatively convergent results if it were to use the
same effective mass resolution. Instead the number of zones inside each star should at least be doubled.}

\begin{figure}
  \centering
  \includegraphics[scale=0.45]{spatial_convergence_rot0}
  \caption{Distance between the two white dwarfs in the unequal mass system, for the first orbit.
           The distance is scaled by the initial orbital distance. 
           We plot at three different resolutions, corresponding to the number of 
           effective zones per dimension in the refined regions.
           \label{fig:unequal_spatial_convergence_inertial}}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[scale=0.45]{spatial_convergence_rot1}
  \caption{Distance between the two white dwarfs in the unequal mass system, for the first orbit.
           The distance is scaled by the initial orbital distance. 
           We plot at three different resolutions, corresponding to the number of 
           effective zones per dimension in the refined regions.
           \label{fig:unequal_spatial_convergence_rotating}}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[scale=0.4]{weak_scaling}
  \caption{\castro\ weak scaling test, performed on Blue Waters at 
    NCSA. Each processor had a fixed amount of work, and we increased the 
    number of simulation zones in concert with the number of processors. The 
    solid curve represents perfect weak scaling, while the blue circles show 
    \castro's performance at each processor count. The vertical axis measures 
    the median time per timestep, normalized to this value for the smallest 
    processor count.\label{fig:weak_scaling}}
\end{figure}

To examine the parallel performance of \castro, we performed both
strong scaling and weak scaling tests on the Blue Waters machine at
the National Center for Supercomputing Applications. For the weak
scaling test, whose results are shown in \autoref{fig:weak_scaling},
we ran a uniform grid binary white dwarf simulation for resolutions of
$128^3$ zones through $2048^3$ zones. The number of processors was
scaled with the number of zones so that each processor had the same
amount of work; the smallest test used 8 processors and the largest
used 32,768 (note that the number of processor cores on a Blue Waters
node is twice the number of floating point units on that node). The
test was run for 10 timesteps, with each timestep including two
Poisson solves and a hydrodynamics update (though for a uniform grid
calculation we generally do not need to perform any multigrid
iterations for the first Poisson solve in a timestep, since the
density distribution has not changed since the end of the last
timestep). We disabled plotfile and checkpoint writing, as well as
calculation of diagnostic information (the latter can contribute to a
significant fraction of the run time at large processor counts if
computed every timestep). We computed the median wall time required
per time step for each simulation, and then normalized this to the
median time per timestep for the smallest simulation. We find
excellent weak scaling through 4,096 processors. At the largest run,
the simulation time required is slightly \textbf{less} than 1.5 times the amount
required for the smallest simulation.  This is due entirely to the
increased cost of the multigrid Poisson solve in each timestep and
this cannot be mitigated except by improving communication or
computation efficiency in the multigrid solver. We observe that this
weak scaling behavior with Poisson gravity is a significant
improvement over the results presented in the first \castro\ paper.

\begin{figure}
  \centering
  \includegraphics[scale=0.4]{strong_scaling}
  \caption{\castro\ strong scaling test performed on the Blue Waters machine at
    NCSA. The vertical axis measures the median time per timestep, and the 
    horizontal axis measures the number of processors in the simulation. Data 
    points are normalized to the time per timestep for the smallest number 
    of processors. The green circles show the data for a simulation with one 
    AMR level (a single uniform grid), the blue diamonds show the data for a simulation with two AMR
    levels (one coarse and one fine), while the red circles show the data 
    for a test with three AMR levels (one coarse and two fine). The fine levels 
    increase the resolution only in the regions around the stars. For each case 
    we draw a solid curve representing perfect strong scaling.\label{fig:strong_scaling}}
\end{figure}

@@

The authors thank John Bell, \textbf{Volker Springel, and Dominic Marcello}
for helpful discussions on gravity 
and hydrodynamics issues. We also thank Noel Scudder and Platon Karpov 
for their help with this project, especially related to visualization of the results, 
and Adam Jacobs for his advice on and assistance with 
running on supercomputing resources. Special thanks are given 
to the organizers of the 2015 Caltech Gravitational Wave 
Astrophysics School, and their supporters: the NSF 
under CAREER award PHY-1151197, the Sherman Fairchild 
Foundation, the LIGO Laboratory at Caltech, and Caltech.

@@

\bibliographystyle{aasjournal}
\input{paper.bbl}
